Variance reduction for Markov chains via martingale representations

### 2020-05-08 aap/Revision/vr-mcmc-revision-final.tex
update

\begin{abstract}
In this paper we propose an efficient variance reduction approach for  additive functionals of Markov chains relying on a novel discrete time martingale representation. Our approach is fully non-asymptotic and does not require the knowledge of the stationary distribution (and even any type of ergodicity) or specific structure of the underlying density.  By rigorously analyzing the  convergence properties of the proposed algorithm, we show that its complexity is indeed asymptotically smaller than one of the naive algorithm. The numerical performance of the new method is illustrated for Markov Chain Monte Carlo (MCMC) methods.
\end{abstract}

@@

One of the popular approaches to variance reduction is the control variates method. We aim at constructing cheaply computable random variable $\zeta$ (control variate) with $\PE[\zeta] = 0$ and $\PE[\zeta^2] < \infty$, such that the variance of the r.v. $f(X) + \zeta$ is small. The complexity of the problem of constructing classes $Z$ of control variates \(\zeta\) satisfying   $\PE[\zeta]=0$ essentially depends on the degree of our knowledge on \(\pi.\)

@@

Unfortunately it is not always possible to generate independent observations distributed according to $\pi$. To overcome this issue one might consider MCMC algorithms, where the exact samples from $\pi$ are replaced by \((X_p),\) \(p=0,1,2,\ldots,\) forming a Markov chain with a marginal distribution of $X_n$ converging to $\pi$ in a suitable metrics as $n$ grows. It is still possible to apply the control variates method in the similar manner to a simple Monte-Carlo case, yet the choice of the optimal control variate becomes much more difficult. Due to significant correlations between the elements of the Markov chain it might be not enough to minimize the marginal variances of \((X_p)_{p \geq 0}\) as it was in independent case. Instead one may choose the control variate by minimizing the corresponding asymptotic variance of the chain as it is suggested in~\cite{belomestny2019esvm}. At the same time it is possible to express the optimal control variate in terms of the solution of the Poisson equation for the corresponding Markov chain \((X_p)\). As it was observed in~\cite{henderson1997variance,henderson2004}, for a time-homogeneous Markov chain \(X_p\) with a stationary distribution \(\pi\), the function $U_{G}(x) := G(x)-\PE[G(X_{1})|X_0 = x]$ has a zero mean with respect to $\pi$ for arbitrary real-valued function \(G: \mathcal{X} \to \rset\), such that \(G  \in L^1(\pi)\). Hence, $U_{G}(x)$ is a valid control functional for a suitable choice of $G$, with the best $G$ given by a solution of the Poisson equation \(\PE [G(X_{1})|X_0 = x]-G(x)=-f(x)+\pi(f)\). For such $G$ we obtain $f(x) + U_G(x) = f(x) - f(x) + \pi(f) = \pi(f)$ leading to an ideal estimator with zero variance. Despite the fact that the Poisson equation involves the quantity of interest \(\pi(f)\)  and can not be  solved explicitly in most cases, this idea still can be used to construct some  approximations for the optimal zero-variance control variates. For example,  \cite{henderson1997variance} proposed to compute approximations for the solution of the Poisson equation for specific Markov chains with particular emphasis on models arising in stochastic network theory. In \cite{dellaportas2012control} and \cite{brosse2018diffusion}  series-type control variates are introduced and studied for reversible Markov chains. It is assumed in \cite{dellaportas2012control}  that the one-step conditional expectations  can be computed explicitly  for a set of basis functions. The authors in \cite{brosse2018diffusion} proposed another approach tailored to diffusion setting which does not require the computation of integrals of basis functions and only involves  applications of the underlying generator.

@@

In this paper we focus on the  Langevin type algorithms which got much attention recently, see \cite{dalalyan2017theoretical,durmus:moulines:2017, MR2353037, MR3861816, MR2977986} and references therein. We propose  a generic variance reduction method for these and other types algorithms, which is purely non-asymptotic and does not require that    the conditional expectations of the corresponding Markov chain can be computed  or that the generator is known analytically. Moreover, we do not need to assume stationarity or/and sampling under the invariant distribution \(\pi.\) We rigorously analyse the convergence of the method and study its complexity. It is shown that  our variance-reduced Langevin algorithm outperforms the standard Langevin algorithms in terms of complexity.

@@

\section{Setup}\label{sec:setup}
Let  \(\mathcal{X}\) be a domain in \( \mathbb{R}^d.\)  Our aim is to numerically compute  expectations of the form
\[
\pi(f)=\int_{\mathcal{X}} f(x)\pi(\rmd x),
\]
where \(f:\) \(\mathcal{X}\longrightarrow \mathbb{R}\) and \(\pi\) is a probability measure supported on \(\mathcal{X}.\)
If  the dimension of the space \(\mathcal{X}\) is large and \(\pi(f)\) can not be computed analytically, one can apply Monte Carlo methods. However, in many practical situations  direct sampling from \(\pi\) is impossible and this precludes the use of plain Monte Carlo methods in this case. One popular alternative to Monte Carlo  is Markov Chain Monte Carlo (MCMC), where one is looking for a discrete time  (possibly non-homogeneous) Markov chain   \((X_p)_{p \in \nset_0}\) such that \(\pi\) is its unique invariant measure. In this paper we study a class of MCMC algorithms with \((X_p)_{p \in \nset_0}\) satisfying the  the following recurrence relation:
\begin{equation}
\label{eq:chain_gen}
X_{p}=\Phi_{p}(X_{p-1},\xi_{p}),\quad p=1,2,\ldots ,\quad X_{0}=x,
\end{equation}
for some i.i.d.  random vectors \(\xi_p\in \mathbb{R}^m\) with distribution \(P_{\xi}\)
and some Borel-measurable
functions $\Phi_{p}\colon\mathcal{X}\times\mathbb{R}^{m}\to\mathcal{X}.$
In fact, this is quite general class of Markov chains (see \cite[Theorem~1.3.6]{moulines2018})
and many well-known MCMC algorithms can be represented in the form \eqref{eq:chain_gen}.
Let us consider two popular examples.
\begin{example}[Unadjusted Langevin Algorithm]
\label{exam:langevin-algorithm}
Fix a sequence of positive time steps \((\gamma_p)_{p\geq 1}.\) Given a Borel function $\mu\colon\mathbb{R}^{d}\to\mathbb{R}^{d}$,
consider a non-homogeneous
discrete-time Markov chain $(X_{p})_{p\geq0}$ defined by
\begin{equation}\label{eq:chain}
X_{p+1}=X_{p}-\gamma_{p+1}\mu(X_{p})+\sqrt{\gamma_{p+1}}Z_{p+1},\end{equation}
where $\left(Z_{p}\right)_{p\geq1}$ is an i.i.d.\ sequence of $d$-dimensional
standard Gaussian random vectors. If
$\mu= \nabla U$
for some continuously differentiable function $U,$ then Markov chain~\eqref{eq:chain} can be used to approximately sample from the density
\begin{equation}\label{eq:stationary_distr}
\pi(x)= Z^{-1} \rme^{-U(x)/2}, \quad Z= \int_{\mathbb{R}^{d}} \rme^{-U(x)/2}\, \rmd x,
\end{equation}
provided that \(Z < \infty\).
This method is usually referred to as Unadjusted Langevin Algorithm (ULA).
Denote by \(W\) the standard \(\mathbb{R}^m\)-valued Brownian motion.
The Markov chain~\eqref{eq:chain}
arises as the Euler-Maruyama discretization
of the Langevin diffusion
\[
\rmd Y_t=-\mu(Y_t)\,\rmd t+ \rmd W_t
\]
with nonnegative time steps $(\gamma_p)_{p\ge1}$,
and,  under mild technical conditions, the latter Langevin diffusion admits $\pi$
of~\eqref{eq:stationary_distr}
as its unique invariant distribution; see \cite{dalalyan2017theoretical} and \cite{durmus:moulines:2017}.
\end{example}
\begin{example}[Metropolis-Adjusted Langevin Algorithm]
The Metropolis-Hastings algorithm
associated with a target density \(\pi\) requires the choice of a sequence of conditional densities  \((q_p)_{p\geq 1}\) also called proposal or candidate kernels. The transition from the value  of the Markov chain \(X_p\)  at time \(p\)
and its value at time \(p + 1\) proceeds via the following transition step:

@@

  Let \((X_t)_{t\in [0,T]}\) be the unique strong solution to a  SDE of the form:

@@

In this section we give a general discrete-time martingale representation for  Markov chains of the type  \eqref{eq:chain_gen} which is  used later to construct an efficient variance reduction algorithm. Let \((\phi_k)_{k\in \mathbb{Z}_+}\) be a complete orthonormal system in \(L^2(\mathbb{R}^m, P_{\xi})\) with \(\phi_0\equiv 1\).  In particular, we have

@@

\begin{cor}
Let \((X^x_{p})_{p\geq 0}\) be a homogeneous Markov chain of the form \eqref{eq:ula_new} with $\Phi_l = \Phi,$ $l\geq 1$. Then for all $p \in \nset,$ $q \leq p$, $j < q \leq p$, $f \in \mathcal{F}_{2,\text{hom}}^p$ and $x \in \rset^d$ it holds in \(L^2(\mathsf{P})\)
\begin{eqnarray*}
f(X^x_{q})=\mathsf{E}\left[\left.f(X^x_{q})\right|\mathcal G_{j}\right]+\sum_{k=1}^{\infty}\sum_{l=j+1}^{q}\bar a_{q-l,k}(X^x_{l-1})\phi_k\left(\xi_{l}\right)
\end{eqnarray*}
where for all $y \in \rset^d$,
\begin{equation}
\label{eq:definition-bar-a-0}
\bar a_{r,k}(y)=\mathsf{E}\left[f(X^y_{r})\phi_k\left(\xi_{1}\right)\right] \,, \text{for $r, k\in \mathbb{N}$.}
\end{equation}
\end{cor}
Another equivalent representation of the coefficients \(a_{p,l,k}\)  turns out to be more useful to construct estimators:
\begin{prop}
Let $q\geq l, k\in \nset$. Then the coefficients \(a_{q,l,k}\) in \eqref{eq:coeff_mart}  can be alternatively represented as
\begin{eqnarray*}
a_{q,l,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{q,l}\left(\Phi_l(x,\xi)\right)\right]
\end{eqnarray*}
with \(Q_{q,l}(x)=\mathsf{E}\left[f(X^x_{l,q})\right],\) \(q\geq l.\)
In the homogeneous case $\Phi_l=\Phi$, the coefficients $\bar{a}_{r,k}$ in \eqref{eq:definition-bar-a-0} are given respectively  for all $l\geq 1,$ by
\begin{equation}
\label{eq:definition-bar-a}
\bar a_{r,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{r}\left(\Phi(x,\xi)\right)\right] \,
\quad \text{with  \(Q_{r}(x)=\mathsf{E}\left[f(X^x_{r})\right],\) $r\in \mathbb{N}.$}
\end{equation}
\end{prop}

@@

$Q_r$ for $r=0,\ldots,n-1,$, defined in \eqref{eq:definition-bar-a}.

@@

If the class of functions $\Psi$ is a linear one of dimension $\mathtt{D}_\Psi,$ then the cost of computing the coefficients $\widehat A_{l,k}$ for all $l=1,\ldots,n,$ and $k=1,\ldots,K,$ is of order  $ \mathtt{D}_\Psi K T^2 n$. Given $(\widehat A_{l,k})$ the cost of computing  $\widehat \pi_{K,n}^N$ is proportional to $\mathtt{D}_\Psi K n.$ At the same time the variance of the standard estimate $ \pi_{n}^N(f)$ is of order $1/(n\gamma)$ and this bound can not be improved, see Lemma~\ref{lem:variance} and Remark~\ref{rem:var-low}. Thus, the ratio of the corresponding cost-variances is of order

@@

Under assumptions {\bf (H1)} and {\bf (H2)} the Markov chain $(X^x_{p})_{p \in \nset_0}$ generated by ULA  is $V-$geometrically ergodic in the sense of \eqref{eq:v_ergodicity} with $V(x) = 1+|x|^2$ and $\rho\in (0,1) $ due to Lemma~\ref{lem:v_ergodicity}. It follows from Lemma~\ref{lem:var_bound}, that
\[
\PVar[\pi_n^N(f)]\lesssim \frac{1}{n(1-\rho^{1/2})} + \left(\frac{\rho^NV(x)+1}{n(1-\rho^{1/2})}\right)^2
\]
It follows from \eqref{eq:rho-ula} that $1-\rho^{1/2} \geq C\gamma$ with a constant $C$ not depending on $\gamma,N$ and $n$, thus yielding an estimate
\[
\PVar[\pi_n^N(f)] \lesssim \frac{1}{n\gamma} + \left(\frac{\rho^NV(x)+1}{n\gamma}\right)^2.
\]

@@

\begin{definition}
\label{def:v_ergodicity}
Let $(X_p)_{p \in \nset_0}$ be a Markov chain taking values in some space $X$ with the Markov kernel $Q$ and stationary distribution $\pi$. We say that $(X_p)_{p \in \nset_0}$ is $V-$geometrically ergodic for a given function $V: X \rightarrow [1; +\infty)$ if there exist real numbers $C > 0$ and $0 < \rho < 1$ such that for any $n \in \nset$,
\begin{equation}
\label{eq:v_ergodicity}
d_V(\delta_xQ^{n},\pi) \leq C\rho^n V(x)
\end{equation}
\end{definition}
In  the sequel, without loss of generality, we assume that $(X_p)_{p \in \nset_0}$ follows ULA dynamics \eqref{eq:chain} with $\nabla U(0)=0$ and  set
\begin{equation}
\label{eq:definition-V}
V(x) = 1 + |x|^2 \eqsp.
\end{equation}
\begin{lem}
  \label{lem:quadratic_behaviour}
  Assume {\bf (H1)} and {\bf (H2)}. Then there exists $K_2 \geq 0$ such that for any $ | x | \geq K_2$ it holds $\ps{\nabla U(x)}{x} \geq (\mathfrak{m}/2) | x |^2$.
\end{lem}
\begin{proof}
 Suppose that the function $\Delta$ in {\bf (H2)} is supported in the ball $\{x:|x|\leq K_1\}$, then we have 
  \begin{align*}
    \ps{\nabla U(x)}{x}
    &= \int_{0}^{K_1/| x |} D^2 U(t x ) [x^{\otimes 2}] \rmd t + \int_{K_1/| x |} ^ 1 D^2 U(t x ) [x^{\otimes 2}] \rmd t\\
    & \geq \mathfrak{m} |x |^2 \{1- K_1 (1 +L/\mathfrak{m})   / | x | \}\eqsp,\quad \mbox{for all} \quad  | x |\geq K_1,
  \end{align*}
which proves the  statement with $K_2=2K_{1}(1+L/\mathfrak{m}).$
\end{proof}
\begin{lem}
\label{lem:drift}
 Assume {\bf (H1)} and {\bf (H2)}.
 Then, for any $\gamma \in (0,\bar{\gamma}]$ with $\overline{\gamma} = \mathfrak{m}/2L^2$  the kernel $Q_\gamma$ from \eqref{eq:ula_kernel} satisfies drift condition
 \begin{equation}
\label{eq:lyapunov}
Q_{\gamma}V(x) \leq \lambda^{\gamma} V(x) + \gamma b
\end{equation}
with   $\lambda = \exp{\left(-\frac{\mathfrak{m}}{2}\right)}$, $b = 2K_2^2(L+\mathfrak{m}) + d + \mathfrak{m}$ with $K_2$ from Lemma~\ref{lem:quadratic_behaviour}, and $\mathfrak{m}$ from {\bf (H2)}.
\end{lem}

@@

\begin{lem}
\label{lem:v_ergodicity}
 Assume  {\bf (H1)} and {\bf (H2)}. Then for $0 < \gamma < \overline{\gamma}= \min(\mathfrak{m} / 2L^2,1/L)$, and for any $x \in \rset^d$  it holds
$$
d_V(\delta_xQ_{\gamma}^{n},\pi_\gamma) \leq C\rho^{\gamma n} V(x)
$$
with $V(x) = 1 + |x|^2$ and some $0 < \rho < 1$ and $C > 0$.
%$$
%C = \left(1 + \exp{\left(-\frac{\mathfrak{m}\gamma}{2}\right)}\right)\left(1+\frac{\overline{b}}{(1-\varepsilon)(1-\exp{\left(-\frac{\mathfrak{m}\gamma}{2}\right)} - \frac{2b}{2+K^2})}\right),
%$$
%
%
%
%$$
%b = \gamma(2K_2^2(L+\mathfrak{m}) + d + \mathfrak{m}), \quad \overline{b} = b\exp{\left(-\frac{\mathfrak{m}\gamma}{2}\right)} + K^2 + 1, 
%$$
%
%\begin{eqnarray*}
%\varepsilon = 2\Phi\left(-K\sqrt{\left(1+\frac{1}{L}\right)\left(1 + 2L\right)}\right),
%\end{eqnarray*}
%
%\begin{eqnarray}
%\label{eq:rho-ula}
%\rho = \exp\left\{-\frac{\gamma m}{4}\frac{\log{(1-\varepsilon)}}{\log{(1-\varepsilon)} + \log{(\exp{\left(-\frac{\mathfrak{m}\gamma}{2}\right)} + \frac{2b}{K^2+2})}}\right\},
%\end{eqnarray}
%where $K_2 > 0$ is from Lemma~\ref{lem:quadratic_behaviour} and $K>0$ is a positive parameter such that 
%$$
%K^2 > 4\frac{2K_2^2(L+m) + d + m}{m}.
%$$
\end{lem}

>>>>

\begin{abstract}
In this paper we propose an efficient variance reduction approach for  additive functionals of Markov chains relying on a novel discrete time martingale representation. Our approach is fully non-asymptotic and does not require the knowledge of the stationary distribution (and even any type of ergodicity) or specific structure of the underlying density.  By rigorously analyzing the  convergence properties of the proposed algorithm, we show that its cost-to-variance relation is indeed  smaller than one of the naive algorithm. The numerical performance of the new method is illustrated for Langevin-type Markov Chain Monte Carlo (MCMC) methods.
\end{abstract}

@@

One of the popular approaches to variance reduction is the control variates method. We aim at constructing cheaply computable random variable $\zeta$ (control variate) with $\PE[\zeta] = 0$ and $\PE[\zeta^2] < \infty$, such that the variance of the r.v. $f(X) + \zeta$ is small. The complexity of the problem of constructing classes of control variates \(\zeta\) satisfying   $\PE[\zeta]=0$ essentially depends on the degree of our knowledge on \(\pi.\)

@@

Unfortunately it is not always possible to generate independent observations distributed according to $\pi$. To overcome this issue one might consider MCMC algorithms, where the exact samples from $\pi$ are replaced by \((X_p),\) \(p=0,1,2,\ldots,\) forming a Markov chain with a marginal distribution of $X_n$ converging to $\pi$ in a suitable metrics as $n$ grows. It is still possible to apply the control variates method in the similar manner to a simple Monte-Carlo case, yet the choice of the optimal control variate becomes much more difficult. Due to significant correlations between the elements of the Markov chain it might be not enough to minimize the marginal variances of \((X_p)_{p \geq 0}\) as it was in independent case. Instead one may choose the control variate by minimizing the corresponding asymptotic variance of the chain as it is suggested in~\cite{belomestny2019esvm}. At the same time it is possible to express the optimal control variate in terms of the solution of the Poisson equation for the corresponding Markov chain \((X_p)\). As it was observed in~\cite{henderson1997variance,henderson2004}, for a time-homogeneous Markov chain \(X_p\) with a stationary distribution \(\pi\), the function $U_{G}(x) := G(x)-\PE[G(X_{1})|X_0 = x]$ has zero mean with respect to $\pi$ for an arbitrary real-valued function \(G: \mathcal{X} \to \rset\), such that \(G  \in L^1(\pi)\). Hence, $U_{G}(x)$ is a valid control functional for a suitable choice of $G$, with the best $G$ given by a solution of the Poisson equation \(\PE [G(X_{1})|X_0 = x]-G(x)=-f(x)+\pi(f)\). For such $G$ we obtain $f(x) + U_G(x) = f(x) - f(x) + \pi(f) = \pi(f)$ leading to an ideal estimator with zero variance. Despite the fact that the Poisson equation involves the quantity of interest \(\pi(f)\)  and can not be  solved explicitly in most cases, this idea still can be used to construct some  approximations for the optimal zero-variance control variates. For example,  \cite{henderson1997variance} proposed to compute approximations for the solution of the Poisson equation for specific Markov chains with particular emphasis on models arising in stochastic network theory. In \cite{dellaportas2012control} and \cite{brosse2018diffusion}  series-type control variates are introduced and studied for reversible Markov chains. It is assumed in \cite{dellaportas2012control}  that the one-step conditional expectations  can be computed explicitly  for a set of basis functions. The authors in \cite{brosse2018diffusion} proposed another approach tailored to diffusion setting which does not require the computation of integrals of basis functions and only involves  applications of the underlying generator.

@@

In this paper we focus on the  Langevin type algorithms which got much attention recently, see \cite{dalalyan2017theoretical,durmus:moulines:2017, MR2353037, MR3861816, MR2977986} and references therein. We propose  a generic variance reduction method for these and other types algorithms, which is purely non-asymptotic and does not require that    the conditional expectations of the corresponding Markov chain can be computed  or that the generator is known analytically. Moreover, we do not need to assume stationarity or/and sampling under the invariant distribution \(\pi.\) We rigorously analyse the convergence of the method and study its complexity. It is shown that  our variance-reduced Langevin algorithm outperforms the standard Langevin algorithms in terms of its cost-to-variance relation.

@@

\section{Setup}\label{sec:setup}
Let  \(\mathcal{X}\) be a domain in \( \mathbb{R}^d.\)  Our aim is to numerically compute  expectations of the form
\[
\pi(f)=\int_{\mathcal{X}} f(x)\pi(\rmd x),
\]
where \(f:\) \(\mathcal{X}\longrightarrow \mathbb{R}\) and \(\pi\) is a probability measure supported on \(\mathcal{X}.\)
If  the dimension of the space \(\mathcal{X}\) is large and \(\pi(f)\) can not be computed analytically, one can apply Monte Carlo methods. However, in many practical situations  direct sampling from \(\pi\) is impossible and this precludes the use of plain Monte Carlo methods in this case. One popular alternative to Monte Carlo  is Markov Chain Monte Carlo (MCMC) where one is looking for a discrete time  (possibly non-homogeneous) Markov chain   \((X_p)_{p \in \nset_0}\) such that \(\pi\) is its unique invariant measure. In this paper we study a class of MCMC algorithms with \((X_p)_{p \in \nset_0}\) satisfying the  the following recurrence relation:
\begin{equation}
\label{eq:chain_gen}
X_{p}=\Phi_{p}(X_{p-1},\xi_{p}),\quad p=1,2,\ldots ,\quad X_{0}=x,
\end{equation}
for some i.i.d.  random vectors \(\xi_p\in \mathbb{R}^m\) with distribution \(P_{\xi}\)
and some Borel-measurable
functions $\Phi_{p}\colon\mathcal{X}\times\mathbb{R}^{m}\to\mathcal{X}.$
In fact, this is quite general class of Markov chains (see \cite[Theorem~1.3.6]{moulines2018})
and many well-known MCMC algorithms can be represented in the form \eqref{eq:chain_gen}.
Let us consider two popular examples.
\begin{example}[Unadjusted Langevin Algorithm]
\label{exam:langevin-algorithm}
Fix a sequence of positive time steps \((\gamma_p)_{p\geq 1}.\) Given a Borel function $\mu\colon\mathbb{R}^{d}\to\mathbb{R}^{d}$,
consider a non-homogeneous
discrete-time Markov chain $(X_{p})_{p\geq0}$ defined by
\begin{equation}\label{eq:chain}
X_{p+1}=X_{p}-\gamma_{p+1}\mu(X_{p})+\sqrt{\gamma_{p+1}}Z_{p+1},\end{equation}
where $\left(Z_{p}\right)_{p\geq1}$ is an i.i.d.\ sequence of $d$-dimensional
standard Gaussian random vectors. If
$\mu= \nabla U$
for some continuously differentiable function $U,$ then Markov chain~\eqref{eq:chain} can be used to approximately sample from the density
\begin{equation}\label{eq:stationary_distr}
\pi(x)= Z^{-1} \rme^{-U(x)/2}, \quad Z= \int_{\mathbb{R}^{d}} \rme^{-U(x)/2}\, \rmd x,
\end{equation}
provided that \(Z < \infty\).
This method is usually referred to as Unadjusted Langevin Algorithm (ULA).
Denote by \(W\) the standard \(\mathbb{R}^m\)-valued Brownian motion.
The Markov chain~\eqref{eq:chain}
arises as the Euler-Maruyama discretization
of the Langevin diffusion
\[
\rmd Y_t=-\mu(Y_t)\,\rmd t+ \rmd W_t
\]
with nonnegative time steps $(\gamma_p)_{p\ge1}$,
and,  under mild technical conditions, the latter Langevin diffusion admits $\pi$
of~\eqref{eq:stationary_distr}
as its unique invariant distribution; see \cite{dalalyan2017theoretical} and \cite{durmus:moulines:2017}.
\end{example}
\begin{example}[Metropolis-Adjusted Langevin Algorithm]
The Metropolis-Hastings algorithm
associated with a target density \(\pi\) requires the choice of a sequence of conditional densities  \((q_p)_{p\geq 1}\) also called proposal or candidate kernels. The transition from the value  of the Markov chain \(X_p\)  at time \(p\)
and its value at time \(p + 1\) proceeds via the following transition step:

@@

  Let \((X_t)_{t\geq 0}\) be the unique strong solution to a  SDE of the form:

@@

In this section we provide a general discrete-time martingale representation for  Markov chains of the type  \eqref{eq:chain_gen} which is  used later to construct an efficient variance reduction algorithm. Let \((\phi_k)_{k\in \mathbb{Z}_+}\) be a complete orthonormal system in \(L^2(\mathbb{R}^m, P_{\xi})\) with \(\phi_0\equiv 1\).  In particular, we have

@@

\begin{cor}
Let \((X^x_{p})_{p\geq 0}\) be a homogeneous Markov chain of the form \eqref{eq:ula_new} with $\Phi_l = \Phi,$ $l\geq 1$. Then for all $p \in \nset,$ $q \leq p$, $j < q \leq p$, $f \in \mathcal{F}_{2,\text{hom}}^p$ and $x \in \rset^d$ it holds in \(L^2(\mathsf{P}),\)
\begin{eqnarray*}
f(X^x_{q})=\mathsf{E}\left[\left.f(X^x_{q})\right|\mathcal G_{j}\right]+\sum_{k=1}^{\infty}\sum_{l=j+1}^{q}\bar a_{q-l,k}(X^x_{l-1})\phi_k\left(\xi_{l}\right)
\end{eqnarray*}
where for all $y \in \rset^d$,
\begin{equation}
\label{eq:definition-bar-a-0}
\bar a_{r,k}(y)=\mathsf{E}\left[f(X^y_{r})\phi_k\left(\xi_{1}\right)\right] \quad  r, k\in \mathbb{N}
\end{equation}
\end{cor}
Another equivalent representation of the coefficients \(a_{p,l,k}\)  turns out to be more useful to construct estimators:
\begin{prop}
Let $q\geq l, k\in \nset$. Then the coefficients \(a_{q,l,k}\) in \eqref{eq:coeff_mart}  can be alternatively represented as
\begin{eqnarray*}
a_{q,l,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{q,l}\left(\Phi_l(x,\xi)\right)\right]
\end{eqnarray*}
with \(Q_{q,l}(x)=\mathsf{E}\left[f(X^x_{l,q})\right],\) \(q\geq l.\)
In the homogeneous case $\Phi_l=\Phi$, the coefficients $\bar{a}_{r,k}$ in \eqref{eq:definition-bar-a-0} are given respectively  for all $l\geq 1,$ by
\begin{equation}
\label{eq:definition-bar-a}
\bar a_{r,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{r}\left(\Phi(x,\xi)\right)\right] \,
\quad \text{with  \(Q_{r}(x)=\mathsf{E}\left[f(X^x_{r})\right],\) $r\in \mathbb{N}.$}
\end{equation}
\end{prop}

@@

$Q_r$ for $r=0,\ldots,n-1,$ defined in \eqref{eq:definition-bar-a}.

@@

If the class of functions $\Psi$ is a linear one of dimension $\mathtt{D}_\Psi,$ then the cost of computing the coefficients $\widehat A_{l,k}$ for all $l=1,\ldots,n,$ and $k=1,\ldots,K,$ is of order  $ \mathtt{D}_\Psi K T^2 n$. Given $(\widehat A_{l,k})$ the cost of computing  $\widehat \pi_{K,n}^N$ is proportional to $\mathtt{D}_\Psi K n.$ At the same time the variance of the standard estimate $ \pi_{n}^N(f)$ is of order $1/(n\gamma)$ and this bound can not be improved, see Lemma~\ref{lem:variance} and Remark~\ref{rem:var-low}. Thus, the ratio of the corresponding cost-to-variance characteristics  is of order

@@

Under assumptions {\bf (H1)} and {\bf (H2)} the Markov chain $(X^x_{p})_{p \in \nset_0}$ is $V-$geometrically ergodic in the sense of \eqref{eq:v_ergodicity} with $V(x) = 1+|x|^2$ and $\rho\in (0,1) $ specified in Lemma~\ref{lem:v_ergodicity}. Due to Lemma~\ref{lem:var_bound},
\[
\PVar[\pi_n^N(f)]\lesssim \frac{1}{n(1-\rho^{1/2})} + \left(\frac{\rho^NV(x)+1}{n(1-\rho^{1/2})}\right)^2
\]
It follows from \eqref{eq:rho-ula} that $1-\rho^{1/2} \geq C\gamma$ with a constant $C$ not depending on $\gamma,N$ and $n$, thus yielding an estimate
\[
\PVar[\pi_n^N(f)] \lesssim \frac{1}{n\gamma} + \left(\frac{\rho^NV(x)+1}{n\gamma}\right)^2.
\]

@@

\begin{definition}
\label{def:v_ergodicity}
Let $(X_p)_{p \in \nset_0}$ be a Markov chain taking values in some space $X$ with the Markov kernel $Q$ and stationary distribution $\pi$. We say that $(X_p)_{p \in \nset_0}$ is $V-$geometrically ergodic for a given function $V: X \rightarrow [1; +\infty)$ if there exist real numbers $C > 0$ and $0 < \rho < 1$ such that for any $n \in \nset$,
\begin{equation}
\label{eq:v_ergodicity}
d_V(\delta_xQ^{n},\pi) \leq C\rho^n V(x)
\end{equation}
\end{definition}
In  the sequel, without loss of generality, we assume that $(X_p)_{p \in \nset_0}$ follows ULA dynamics \eqref{eq:chain} with $\nabla U(0)=0$ and  set
\begin{equation}
\label{eq:definition-V}
V(x) = 1 + |x|^2 \eqsp.
\end{equation}
\begin{lem}
  \label{lem:quadratic_behaviour}
  Assume {\bf (H1)} and {\bf (H2)}. Then there exists $K_2 \geq 0$ such that for any $ | x | \geq K_2$ it holds $\ps{\nabla U(x)}{x} \geq (\mathfrak{m}/2) | x |^2$.
\end{lem}
\begin{proof}
 Suppose that the function $\Delta$ in {\bf (H2)} is supported in the ball $\{x:|x|\leq K_1\}$, then we have 
  \begin{align*}
    \ps{\nabla U(x)}{x}
    &= \int_{0}^{K_1/| x |} D^2 U(t x ) [x^{\otimes 2}] \rmd t + \int_{K_1/| x |} ^ 1 D^2 U(t x ) [x^{\otimes 2}] \rmd t\\
    & \geq \mathfrak{m} |x |^2 \{1- K_1 (1 +L/\mathfrak{m})   / | x | \}\eqsp,\quad \mbox{for all} \quad  | x |\geq K_1,
  \end{align*}
which proves the  statement with $K_2=2K_{1}(1+L/\mathfrak{m}).$
\end{proof}
\begin{lem}
\label{lem:drift}
 Assume {\bf (H1)} and {\bf (H2)}.
 Then, for any $\gamma \in (0,\bar{\gamma}]$ with $\overline{\gamma} = \mathfrak{m}/2L^2$  the kernel $Q_\gamma$ from \eqref{eq:ula_kernel} satisfies drift condition
 \begin{equation}
\label{eq:lyapunov}
Q_{\gamma}V(x) \leq \rho^{\gamma} V(x) + \gamma C
\end{equation}
with   $\rho = \exp{\left(-\frac{\mathfrak{m}}{2}\right)}$, $C = 2K_2^2(L+\mathfrak{m}) + d + \mathfrak{m}$ with $K_2$ from Lemma~\ref{lem:quadratic_behaviour}, and $\mathfrak{m}$ from {\bf (H2)}.
\end{lem}

@@

\begin{lem}
\label{lem:v_ergodicity}
 Assume  {\bf (H1)} and {\bf (H2)}. Then for $0 < \gamma < \overline{\gamma}= \min(\mathfrak{m} / 4L^2,1/L)$, and for any $x \in \rset^d$  it holds
$$
d_V(\delta_xQ_{\gamma}^{n},\pi_\gamma) \leq C\rho^{ n}\left(V(x) + \pi_\gamma(V)\right)
$$
with $V(x) = 1 + |x|^2$ and constants
$$
C = \left(1 + \exp{\left(-\frac{\mathfrak{m}\gamma}{2}\right)}\right)\left(1+\frac{\overline{b}}{(1-\varepsilon)(1-\exp{\left(-\frac{\mathfrak{m}\gamma}{2}\right)} - \frac{2b}{2+K^2})}\right),
$$

$$
b = \gamma(2K_2^2(L+\mathfrak{m}) + d + \mathfrak{m}), \quad \overline{b} = b\exp{\left(-\frac{\mathfrak{m}\gamma}{2}\right)} + K^2 + 1, 
$$

\begin{eqnarray*}
\varepsilon = 2\Phi\left(-K\sqrt{\left(1+\frac{1}{L}\right)\left(1 + 2L\right)}\right),
\end{eqnarray*}

\begin{eqnarray}
\label{eq:rho-ula}
\rho = \exp\left\{-\frac{\gamma m}{4}\frac{\log{(1-\varepsilon)}}{\log{(1-\varepsilon)} + \log{(\exp{\left(-\frac{\mathfrak{m}\gamma}{2}\right)} + \frac{2b}{K^2+2})}}\right\},
\end{eqnarray}
where $K_2 > 0$ is from Lemma~\ref{lem:quadratic_behaviour} and $K>0$ is a positive parameter such that 
$$
K^2 > 4\frac{2K_2^2(L+m) + d + m}{m}.
$$
\end{lem}

\begin{proof} Note that the condition {\bf (H1)} implies that for any $x,y \in \rset^d$,
\[
|x-y-\gamma(\nabla U(x) - \nabla U(y))|^2 \leq (1 + 2L\gamma + \gamma^2)|x-y|^2
\] 
Hence,   \cite[Corollary 5]{DurmusBortoli} implies that
\[
\|\delta_xQ^{[1/\gamma]}_{\gamma} - \delta_yQ^{[1/\gamma]}_{\gamma}\|_{\text{TV}} \leq 2(1-\varepsilon)
\]
with $\varepsilon = 2\Phi\left(-K\sqrt{\left(1+\frac{1}{L}\right)\left(1 + 2L\right)}\right)$ for all $x,y$ such that $|x|,|y| \leq K$, which means that the kernel $Q_{\gamma}$ satisfies $([1/\gamma],\varepsilon)$-Doeblin condition on the set $\{x: V(x) < 1+K^2\}$. Together with drift condition \eqref{eq:lyapunov} it allows to apply \cite[Theorem~19.4.1]{moulines2018} with appropriate constants.
\end{proof}
